Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-03-31 paper/manuscript.tex

\subsection*{Our measurements can help distinguish between viral clades that reach low and high frequencies in nature}
Mutations occurring in the H3N2 population have widely varying evolutionary outcomes~\ref{fig:frequency_trajectory}.
Some mutations appear, spread and fix in the population, while others briefly circulate before dieing out.
We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
If a mutation is deleterious, we would expect it to quickly die out, whereas beneficial mutations are more likely to expand in frequency.
Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations separately.
As such, each mutation is shown as a separate circle occuring on a separate branch in Figure \ref{fig:frequency_trajectory}.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.png}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of viral clades and their relation to the effects of individual mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
Circles indicate amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The bottom panel shows the frequency trajectory of the clade downstream of each mutation, with clades colored according to the mutational effects as in the top panel.
It is obvious that most clades that reach high frequency have mutations that are measured to be relatively favorable in our experiments.
\comment{John: Can you update figure? We defer to you and Trevor on choice of colors, I know there was some Slack discussion of this but can't remember answer. But definitely let's increase the size of the legend and make it $<-10$, and $> 0$ rather than having the \emph{inf} in the labels. Also, can you make y-axis at bottom ``clade frequency'' rather than ``mutation frequency'' as we use that terminology in the text instead.}
}
\end{figure}

Here, we examine to what degree our experimental measurements correlate with realized evolutionary outcomes.
Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
To analyze the relationship between mutational effect and clade frequency more quantitatively, we next partitioned the phylogeny of human H3N2 influenza HA into sequences that split from the line of descent (trunk of the phylogenetic tree) before or after the Perth/2009 strain used in our experiments (see Figure~\ref{suppfig:tree}).
The rationale for partitioning into pre- and post-Perth/2009 sequences is that we wanted to test the utility of our measurements for prospective as well as post-hoc analyses.
\comment{John: do you have a sentence to add about how we used max tips to avoid passaging artifacts?}
Figure~\ref{fig:muteffect_maxfreq} shows that there is a statistically significant trend for clades that reach high frequency to have more favorable mutations than clades that quickly die out.
Importantly, the effect size is similar for both the pre- and post-Perth/2009 sequence groups.

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf The experimental measurements are informative about the evolutionary fate of viral clades.}
(A) The correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by viral clades.
The plots show the Spearman correlation $\rho$ and a $P$-value obtained by repeatedly randomizing the experimental measurements among sites \comment{John: is this right about how the $P$-value was calculated?}.
(B) The distribution of mutational effects among clades binned by the maximum frequency that they reach.
Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
\comment{John: label the top and bottom into A and B as indicated in the legend above, adjust font sizes (particularly, tick labels too big), adjust capitalization to be consistent with other figures (which usually start lower-case unless proper noun), increasing spacing between A and B panels. Also, no need for a title at top and figure can probably get smaller. Some of this might be best done in Inkscape working with Juhye.}
}
\end{figure*}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.75\textwidth]{figs/muteffect_by_maxfreq_WSN_2.pdf}}
\caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
{\bf The distribution of mutational effects measured in H1 HA among H3N2 clades binned by the maximum frequency that they reach.}
This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}B here so that they look consistent.}
}
\end{suppfigure}

>>>>

\subsection*{Our measurements can help distinguish between viral clades that reach low and high frequencies in nature}
Mutations occurring in the H3N2 population have widely varying evolutionary outcomes~\ref{fig:frequency_trajectory}.
Some mutations appear, spread and fix in the population, while others briefly circulate before dieing out.
We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
If a mutation is deleterious, we would expect it to quickly die out, whereas beneficial mutations are more likely to expand in frequency.
We anticipated that clades with deleterious mutations represented by low mutational effects would never reach high global frequencies. 
Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations separately.
As such, each mutation is shown as a separate circle occuring on a separate branch in Figure \ref{fig:frequency_trajectory}.
We identified all mutations on a phylogeny branch and assigned a single mutational effect value based on the sum of effects for each mutation on the branch.

In contrast, we expected clades with beneficial mutations represented by high mutational effects to dominate globally and fix. 
Correspondingly, we observed few clades in our phylogeny with strongly deleterious mutational effects and rarely observed these clades reaching high frequency (Figure~\ref{fig:frequency_trajectory}). 
Clades that reached high global frequencies generally had neutral or beneficial mutational effects.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.png}
\caption{\label{fig:frequency_trajectory}
{\bf Frequency trajectories of viral clades and their relation to the effects of individual mutations.}
The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
Circles indicate amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
The bottom panel shows the frequency trajectory of the clade downstream of each mutation, with clades colored according to the mutational effects as in the top panel.
It is obvious that most clades that reach high frequency have mutations that are measured to be relatively favorable in our experiments.
}
\end{figure}

We further quantified this relationship by calculating a Spearman rank correlation between mutational effects and maximum global frequencies of clades in the H3N2 phylogeny.
Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
To minimize effects related to the genetic background of the DMS strain, we excluded clades closely related to the DMS strain itself and partitioned the remaining clades into those pre- and post-dating the Perth/2009 strain.
We additionally excluded clades from the post-Perth partition that were sampled in 2014 or after and that have not had enough evolutionary time to reach fixation.
We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
In both pre-Perth and post-Perth clades, we found a modest, statistically significant relationship between maximum clade frequency and mutational effect (Figure~\ref{fig:muteffect_maxfreq}A).
The similar effect size for both partitions suggests that our experimental measurements can be used to predict clade frequencies despite the genetic background of the DMS strain.

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf The experimental measurements are informative about the evolutionary fate of viral clades.}
(A) The correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by viral clades.
The plots show the Spearman correlation $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which the Spearman correlation $\rho$ between maximum frequency and randomized mutational effects was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects among clades binned by the maximum frequency that they reach.
Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{figure*}

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.75\textwidth]{figs/muteffect_by_maxfreq_WSN_2.pdf}}
\caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
{\bf The distribution of mutational effects measured in H1 HA among H3N2 clades binned by the maximum frequency that they reach.}
This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
}
\end{suppfigure}
